---
title: "AACR GENIE `r params$release` Data Guide"
format:
  pdf:
    toc: true
    number-sections: false
    colorlinks: true
editor: source
date: today
author: "AACR"
params:
  release: "13.6-consortium"
  db_mapping_synid: "syn10967259"
---

```{r setup, echo=F, include=F}
library(synapser)
library(glue)
# library(xtable)
library(knitr)
library(ggplot2)
library(ggpubr)
synLogin()


get_release_folder_synid <- function(database_synid_mappingid, release) {
  database_synid_mapping = synTableQuery(glue('select * from {synid}',
                                              synid = database_synid_mappingid))
  database_synid_mappingdf = synapser::as.data.frame(database_synid_mapping)
  release_folder_ind = database_synid_mappingdf$Database == "releaseFolder"
  release_folder_fileview_synid = database_synid_mappingdf$Id[release_folder_ind]

  choose_from_release = synTableQuery(glue("select distinct(name) as releases from {synid} where ",
                                           "name not like 'Release%' and name <> 'case_lists'",
                                           synid = release_folder_fileview_synid))
  releases = synapser::as.data.frame(choose_from_release)
  if (!any(releases$releases %in% release)) {
    stop(glue("Must choose correct release: {releases}",
              releases = paste0(releases$releases, collapse = ", ")))
  }

  release_folder = synTableQuery(glue("select id from {synid} where name = '{release}'",
                                      synid = release_folder_fileview_synid,
                                      release = release),
                                 includeRowIdAndRowVersion = F)
  release_folder$asDataFrame()$id
}


get_file_mapping = function(release_folder_synid) {
  release_ent = synGet(release_folder_synid)
  print(release_ent$properties$name)
  release_files = synGetChildren(release_folder_synid)
  release_files_list = as.list(release_files)
  file_mapping = sapply(release_files_list, function(release_file) {
    mapping = c()
    if (release_file$name == "genie_combined.bed") {
      release_file$name = "genomic_information.txt"
    }
    mapping[release_file$name] = release_file$id
    mapping
  })
  file_mapping
}


get_list_assay_info_table <- function(allowed_values, assayinfodf, col) {
  na_ind = is.na(assayinfodf[,col])
  seq_assays = assayinfodf$SEQ_PIPELINE_ID[!na_ind]
  col_per_panel = matrix(nrow = length(seq_assays),
                         ncol = length(allowed_values),
                         dimnames = list(seq_assays, allowed_values))
  for (panel in seq_assays) {
    str_value = assayinfodf[,col][assayinfodf$SEQ_PIPELINE_ID == panel]
    str_value_vector = unlist(strsplit(str_value, ";"))
    col_per_panel[panel, ] = ''
    col_per_panel[panel, str_value_vector] = "\\checkmark"
  }
  col_per_panel
}

database_synid_mappingid = params$db_mapping_synid
release = params$release

release_folder_synid <- get_release_folder_synid(database_synid_mappingid, release)

# Get release files mapping
release_files_mapping = get_file_mapping(release_folder_synid)

```

# About this Document

This document provides an overview of `r params$release` release of American Association for Cancer Research (AACR) GENIE data.

# Version of Data

AACR GENIE Project Data: Version `r params$release`

AACR Project GENIE data versions follow a numbering scheme derived from [semantic versioning](https://semver.org/) where the digits in the version correspond to: major.patch-release-type. "Major" releases are public releases of new sample data. "Patch" releases are corrections to major releases, including data retractions. "Releasetype" refers to whether the release is a public AACR Project GENIE release or a private/consortium-only release. Public releases will be denoted with the nomenclature "X.X-public" and consortium-only private releases will be denoted with the nomenclature "X.X-consortium".

# Data Access

AACR GENIE Data is currently available via two mechanisms:

-   Synapse Platform (Sage Bionetworks): https://synapse.org/genie

-   cBioPortal for Cancer Genomics (MSK): https://www.cbioportal.org/genie/

# Terms Of Access

All users of the AACR Project GENIE data must agree to the following terms of use; failure to abide by any term herein will result in revocation of access.

-   Users will not attempt to identify or contact individual participants from whom these data were collected by any means.

-   Users will not redistribute the data without express written permission from the AACR Project GENIE Coordinating Center (send email to: info\@aacrgenie.org).

When publishing or presenting work using or referencing the AACR Project GENIE dataset please include the following attributions:

-   Please cite: *The AACR Project GENIE Consortium. AACR Project GENIE: Powering Precision Medicine Through An International Consortium, Cancer Discov. 2017 Aug;7(8):818-831* and include the version of the dataset used.

-   The authors would like to acknowledge the American Association for Cancer Research and its financial and material support in the development of the AACR Project GENIE registry, as well as members of the consortium for their commitment to data sharing. Interpretations are the responsibility of study authors.

Posters and presentations should include the AACR Project GENIE logo.

# Introduction to AACR GENIE

The AACR Project Genomics, Evidence, Neoplasia, Information, Exchange (GENIE) is a multi-phase, multi-year, international data-sharing project that aims to catalyze precision cancer medicine. The GENIE platform will integrate and link clinical-grade cancer genomic data with clinical outcome data for tens of thousands of cancer patients treated at multiple international institutions. The project fulfills an unmet need in oncology by providing the statistical power necessary to improve clinical decision-making, to identify novel therapeutic targets, to understand of patient response to therapy, and to design new biomarker-driven clinical trials. The project will also serve as a prototype for aggregating, harmonizing, and sharing clinical-grade, next-generation sequencing (NGS) data obtained during routine medical practice.

The data within GENIE is being shared with the global research community. The database currently contains CLIA-/ISO-certified genomic data obtained during the course of routine practice at multiple international institutions (Table 1), and will continue to grow as more patients are treated at additional participating centers.

```{r, echo=F, include=F}
assay_synid = release_files_mapping[['assay_information.txt']]
assayinfo = suppressMessages(synGet(assay_synid, followLink = T))
assayinfodf = read.csv(assayinfo$path, sep = "\t", stringsAsFactors = F)
processed_centers = unique(assayinfodf$CENTER)
centers = suppressMessages(synTableQuery(glue('select "Center Abbreviation", "Center" from syn16982837 ',
                                              "where \"Center Abbreviation\" in ('{centers}')",
                                              centers = paste(processed_centers, collapse = "','")),
                                         includeRowIdAndRowVersion = F))
```

```{r, echo=F}
#| tbl-cap: "Participating Centers"

kable(centers$asDataFrame())
```

# Human Subjects Protection and Privacy

Protection of patient privacy is paramount, and the AACR GENIE Project therefore requires that each participating center share data in a manner consistent with patient consent and center-specific Institutional Review Board (IRB) policies. The exact approach varies by center, but largely falls into one of three categories: IRB-approved patient-consent to sharing of de-identified data, captured at time of molecular testing; IRB waivers and; and IRB approvals of GENIE-specific research proposals. Additionally, all data has been de-identified via the HIPAA Safe Harbor Method. Full details regarding the HIPAA Safe Harbor Method are available online at: https://www.hhs.gov/hipaa/for-professionals/privacy/special-topics/de-identification/.

# Summary of Sequence Pipeline

Traditionally, the SEQ_ASSAY_ID was used as an institution's identifier for their assays when each assay had one associated gene panel. As GENIE grew, we wanted to support an assay having multiple gene panels. SEQ_ASSAY_ID was repurposed to be an identifier for a center's assay OR panel. For those centers that have multiple panels per assay, we introduced SEQ_PIPELINE_ID (pipeline), which encompasses multiple SEQ_ASSAY_ID (panel).

```{r, echo=F, include=F}
pipelinedf = assayinfodf[!duplicated(assayinfodf$SEQ_PIPELINE_ID), ]
processed_centers = unique(assayinfodf$CENTER)
```

```{r, echo=F}
#| tbl-cap: "Number of pipelines per Center"

kable(table(pipelinedf$CENTER), col.names = c("Center", "# of Pipelines"))
```

```{r, echo=F}
library_selection <- ggplot(pipelinedf, aes(library_selection)) +
  geom_bar() +
  theme_bw() +
  labs(x = "Library Selection", y = "# of Pipelines")
library_strategy <- ggplot(pipelinedf, aes(library_strategy)) +
  geom_bar() +
  theme_bw() +
  labs(x = "Library Strategy", y = "# of Pipelines")

platform <- ggplot(pipelinedf, aes(platform)) +
  geom_bar() +
  theme_bw() +
  labs(x = "Platform", y = "# of Pipelines")

specimen_tumor_cellularity <- ggplot(pipelinedf,
                                     aes(specimen_tumor_cellularity)) +
  geom_bar() +
  theme_bw() +
  labs(x = "Specimen Tumor Cellularity", y = "# of Pipelines")

h = ggarrange(library_selection, library_strategy,
              platform,
              specimen_tumor_cellularity,
              labels = c("A", "B", "C", "D"),
              ncol = 2, nrow = 2)
print(h)
```

# Genomic Profiling at Each Center

```{r}
#| eval: true
#| echo: false
#| results: asis

for (center in sort(processed_centers)) {
  center = tolower(center)
  a <- knitr::knit_child(glue::glue('genomic_profiles/{center}.qmd'), quiet=T)
  cat(a, sep="\n")
}

```

# Pipeline for Annotating Mutations and Filtering Putative Germline SNPs

Contributing GENIE centers provided mutation data in [Variant Call Format (VCF)](http://samtools.github.io/hts-specs/) or Mutation Annotation Format ([GDC MAF v1.0.0](https://docs.gdc.cancer.gov/Data/File_Formats/MAF_Format/)) with additional fields for read counts supporting variant alleles, reference alleles, and total depth. Some "MAF-like" text files with [minimal required columns](https://github.com/mskcc/vcf2maf/blob/v1.6.17/data/minimalist_test_maf.tsv) were also received from the participating centers. These various input formats were converted into a complete tab-separated MAF format, with [Genome Nexus](https://www.genomenexus.org/)

While the GENIE data available from Sage contains all mutation data, the following mutation types are automatically filtered upon import into the [cBioPortal](http://www.cbioportal.org/genie): Silent, Intronic, 3' UTR, 3' Flank, 5' UTR, 5' Flank and Intergenic region (IGR).

Seventeen of the nineteen GENIE participating centers performed tumor-only sequencing i.e. without also sequencing a patient-matched control sample like blood, to isolate somatic events. These centers minimized artifacts and germline events using pooled controls from unrelated individuals, or using databases of known artifacts, common germline variants, and recurrent somatic mutations. However, there remains a risk that such centers may inadvertently release germline variants that can theoretically be used for patient re-identification. To minimize this risk, the GENIE consortium developed a stringent germline filtering pipeline, and applied it uniformly to all variants across all centers. Genome Nexus will report gnomAD AFs by querying the Genome Aggregation Database to determine common variants. If any variant has a max gnomAD AF from any subpopulation that is over 0.0005, the variant will be filtered out. This cutoff was selected because it tagged no more than 1% of the somatic calls across all MSK-IMPACT samples with patient-matched controls.

-   Hotspots from Chang et al. minus some likely artifacts.(dx.doi.org/10.1038/nbt.3391

-   Somatic mutations associated with clonal hematopoietic expansion from Xie et al. (dx.doi.org/10.1038/nm.3733)

-   Somatically mutable germline sites at MSH6:F1088, TP53:R290, TERT:E280, ASXL1:G645_G646.

# Description of Data Files

Description on most of the data files can be found under the "Data files" section for [cBioPortal file formats](https://docs.cbioportal.org/5.1-data-loading/data-loading/file-formats).

+------------------------------------------------------+------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| File Name                                            | Description                                                                  | Details                                                                                                                                                                                                                                         |
+======================================================+==============================================================================+=================================================================================================================================================================================================================================================+
| data_mutations_extended.txt                          | Mutation data                                                                | [MAF format](https://docs.cbioportal.org/5.1-data-loading/data-loading/file-formats#mutation-data)                                                                                                                                              |
+------------------------------------------------------+------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| data_CNA.txt                                         | Discritized copy number data.                                                | [CNA format](https://docs.cbioportal.org/5.1-data-loading/data-loading/file-formats#discrete-copy-number-data)                                                                                                                                  |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      | Note: Not all centers contributed copy number data to GENIE                  |                                                                                                                                                                                                                                                 |
+------------------------------------------------------+------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| data_sv.txt                                          | Structural variant data                                                      | [Structural Variant format](https://docs.cbioportal.org/file-formats/#structural-variant-data)                                                                                                                                                  |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      | Note: not all centers contributed structural rearrangement data to GENIE     |                                                                                                                                                                                                                                                 |
+------------------------------------------------------+------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| genomic_information.txt                              | File describing genomic coordinates covered by all platforms for GENIE data. | This is not a cBioPortal file format.                                                                                                                                                                                                           |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      | Note: Not all centers contributed copy number data to GENIE.                 | -   Chromosome, Start_Position, End_Position: Gene positions                                                                                                                                                                                    |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   Hugo_Symbol: Re-mapped gene symbol based on gene positions.                                                                                                                                                                                 |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   ID: Center submitted gene symbols.                                                                                                                                                                                                          |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   SEQ_ASSAY_ID: The institutional assay identifier for genomic testing platform.                                                                                                                                                              |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   Feature_Type: "exon", "intron", or "intergenic".                                                                                                                                                                                            |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   includeInPanel: Used to define gene panel files for cBioPortal.                                                                                                                                                                             |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   clinicalReported: These are the genes that were clinically Reported. Blank means information not provided.                                                                                                                                  |
+------------------------------------------------------+------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| assay_information.txt                                | Assay information                                                            | This is not a cBioPortal file format.                                                                                                                                                                                                           |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   is_paired_end, library_selection, library_strategy, platform, read_length, target_capture_kit, instrument_model: defined by [GDC read group](https://docs.gdc.cancer.gov/Data_Dictionary/viewer/#?view=table-definition-view&id=read_group) |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   number_of_genes: Number of genes from which variants are called                                                                                                                                                                             |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   variant_classifications: List of types of variants that are reported for this assay.                                                                                                                                                        |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   gene_padding: Number of base pairs to add to exon endpoints for the inBED filter.                                                                                                                                                           |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   alteration_types: List of alteration types.                                                                                                                                                                                                 |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   specimen_type: List of specimen types                                                                                                                                                                                                       |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   specimen_tumor_cellularity: Tumor Cellularity Cutoff                                                                                                                                                                                        |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   calling_strategy: Tumor only or tumor normal                                                                                                                                                                                                |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | -   coverage: List of coverage                                                                                                                                                                                                                  |
+------------------------------------------------------+------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| data_cna_hg19.seg                                    | Segmented copy number data.                                                  | [SEG format](https://docs.cbioportal.org/5.1-data-loading/data-loading/file-formats#segmented-data)                                                                                                                                             |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      | Note: Not all centers contributed copy number data to GENIE.                 |                                                                                                                                                                                                                                                 |
+------------------------------------------------------+------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| data_clinical_patient.txt / data_clinical_sample.txt | De-identified clinical data.                                                 | [Clinical format](https://docs.cbioportal.org/5.1-data-loading/data-loading/file-formats#clinical-data)                                                                                                                                         |
|                                                      |                                                                              |                                                                                                                                                                                                                                                 |
|                                                      |                                                                              | See Clinical Data section below for more details.                                                                                                                                                                                               |
+------------------------------------------------------+------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

: GENIE Data Files {tbl-colwidths="\[35,35,30\]"}

# Clinical Data

+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Data Element          | Example Values                | Data Description                                                                                                                                                                                                                                                                                                                                                                                                                      |
+=======================+===============================+=======================================================================================================================================================================================================================================================================================================================================================================================================================================+
| AGE_AT_SEQ_REPORT     | Integer values, \<18, or \>89 | The age of the patient at the time                                                                                                                                                                                                                                                                                                                                                                                                    |
|                       |                               | that the sequencing results were reported. Age is masked for patients aged 90 years and greater and for patients under 18 years.                                                                                                                                                                                                                                                                                                      |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| CENTER                | MSK                           | The center submitting the clinical and genomic data                                                                                                                                                                                                                                                                                                                                                                                   |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ETHNICITY             | Non-Spanish/nonHispanic       | Indication of Spanish/Hispanic                                                                                                                                                                                                                                                                                                                                                                                                        |
|                       | Spanish/Hispanic              | origin of the patient; this data                                                                                                                                                                                                                                                                                                                                                                                                      |
|                       | Unknown                       | element maps to the NAACCR                                                                                                                                                                                                                                                                                                                                                                                                            |
|                       |                               | v16, Element #190. Institutions                                                                                                                                                                                                                                                                                                                                                                                                       |
|                       |                               | not collecting Spanish/Hispanic                                                                                                                                                                                                                                                                                                                                                                                                       |
|                       |                               | origin have set this column to                                                                                                                                                                                                                                                                                                                                                                                                        |
|                       |                               | Unknown.                                                                                                                                                                                                                                                                                                                                                                                                                              |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ONCOTREE_CODE         | LUAD                          | The primary cancer diagnosis                                                                                                                                                                                                                                                                                                                                                                                                          |
|                       |                               | "main type", based on the OncoTree ontology. The version                                                                                                                                                                                                                                                                                                                                                                              |
|                       |                               | of Oncotree ontology that was                                                                                                                                                                                                                                                                                                                                                                                                         |
|                       |                               | used for GENIE 13.0-public is                                                                                                                                                                                                                                                                                                                                                                                                         |
|                       |                               | oncotree_2021_11_02                                                                                                                                                                                                                                                                                                                                                                                                                   |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| PATIENT_ID            | GENIE-JHU-1234                | The unique, anonymized patient identifier for the GENIE project. Conforms to the following the convention: GENIE-CENTER-1234. The first component is the string, "GENIE"; the second component is the Center abbreviation. The third component is an anonymized unique identifier for the patient.                                                                                                                                    |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| PRIMARY_RACE          | Asian                         | The primary race recorded for the patient; this data element maps to the NAACCR v16, Element #160. For institutions collecting more than one race category, this race code is the primary race for the patient. Institutions not collecting race have set this field to Unknown.                                                                                                                                                      |
|                       | Black                         |                                                                                                                                                                                                                                                                                                                                                                                                                                       |
|                       | Native American               |                                                                                                                                                                                                                                                                                                                                                                                                                                       |
|                       | Other                         |                                                                                                                                                                                                                                                                                                                                                                                                                                       |
|                       | Unknown                       |                                                                                                                                                                                                                                                                                                                                                                                                                                       |
|                       | White                         |                                                                                                                                                                                                                                                                                                                                                                                                                                       |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| SAMPLE_ID             | GENIE-JHU-1234-9876           | The unique, anonymized sample identifier for the GENIE project. Conforms to the following the convention: GENIE-CENTER-1234-9876. The first component is the string, "GENIE"; the second component is the Center abbreviation. The third component is an anonymized, unique patient identifier. The fourth component is a unique identifier for the sample that will distinguish between two or more specimens from a single patient. |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| SAMPLE_TYPE           | Primary                       | The specimen's type including                                                                                                                                                                                                                                                                                                                                                                                                         |
|                       | Metastasis                    | primary, metastasis and etc\...                                                                                                                                                                                                                                                                                                                                                                                                       |
|                       | Unspecified                   |                                                                                                                                                                                                                                                                                                                                                                                                                                       |
|                       | Not Applicable or Heme        |                                                                                                                                                                                                                                                                                                                                                                                                                                       |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| SAMPLE_TYPE_DETAILED  | Primary tumor                 | The specimen's detailed type                                                                                                                                                                                                                                                                                                                                                                                                          |
|                       | Lymph node metastasis         | based on its location, including                                                                                                                                                                                                                                                                                                                                                                                                      |
|                       |                               | primary site, site of local recurrence, distant metastasis or                                                                                                                                                                                                                                                                                                                                                                         |
|                       |                               | hematologic malignancy.                                                                                                                                                                                                                                                                                                                                                                                                               |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| SEQ_ASSAY_ID          | DFCI-ONCOPANEL-1              | The institutional assay identifier for genomic testing platform. Components are separated by hyphens, with the first component corresponding to the Center's abbreviation. All specimens tested by the same platform should have the same identifier.                                                                                                                                                                                 |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| SEX                   | Female\                       | The patient's sex code; this data                                                                                                                                                                                                                                                                                                                                                                                                     |
|                       | Male                          | element maps to the NAACCR                                                                                                                                                                                                                                                                                                                                                                                                            |
|                       |                               | v16, Element #220.                                                                                                                                                                                                                                                                                                                                                                                                                    |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| CANCER_TYPE           | Non-Small Cell Lung Cancer    | The primary cancer diagnosis                                                                                                                                                                                                                                                                                                                                                                                                          |
|                       |                               | "main type", based on the OncoTree ontology. For example,                                                                                                                                                                                                                                                                                                                                                                             |
|                       |                               | the OncoTree code of LUAD                                                                                                                                                                                                                                                                                                                                                                                                             |
|                       |                               | maps to: "Non-Small Cell Lung                                                                                                                                                                                                                                                                                                                                                                                                         |
|                       |                               | Cancer". The version of Oncotree ontology that was used for GENIE 13.0-public is oncotree_2021_11_02.                                                                                                                                                                                                                                                                                                                                 |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| CANCER_TYPE_DETAILED  | Lung Adenocarcinoma           | The primary cancer diagnosis label, based on the OncoTree ontology. For example, the OncoTree code of LUAD maps to the label: "Lung Adenocarcinoma (LUAD)". The version of Oncotree ontology that was used for GENIE 13.0-public is oncotree_2021_11_02                                                                                                                                                                               |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| BIRTH_YEAR            | 1965                          | The four-digit year associated                                                                                                                                                                                                                                                                                                                                                                                                        |
|                       |                               | with the patient's date of birth.                                                                                                                                                                                                                                                                                                                                                                                                     |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| YEAR_CONTACT          | 1990                          | Record the year that the patient                                                                                                                                                                                                                                                                                                                                                                                                      |
|                       |                               | is last known to be alive - data                                                                                                                                                                                                                                                                                                                                                                                                      |
|                       |                               | to be derived from information                                                                                                                                                                                                                                                                                                                                                                                                        |
|                       |                               | including and EHR, tumor registry, or other system.                                                                                                                                                                                                                                                                                                                                                                                   |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| YEAR_DEATH            | 2010                          | Year of death.                                                                                                                                                                                                                                                                                                                                                                                                                        |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| INT_CONTACT           | 15000                         | Interval in days from DOB to                                                                                                                                                                                                                                                                                                                                                                                                          |
|                       |                               | date of last contact.                                                                                                                                                                                                                                                                                                                                                                                                                 |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| INT_DOD               | 15000                         | Interval in days from DOB to                                                                                                                                                                                                                                                                                                                                                                                                          |
|                       |                               | DOD.                                                                                                                                                                                                                                                                                                                                                                                                                                  |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| DEAD                  | Yes                           | Is this patient known to be deceased?                                                                                                                                                                                                                                                                                                                                                                                                 |
|                       | No                            |                                                                                                                                                                                                                                                                                                                                                                                                                                       |
+-----------------------+-------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

: GENIE Clinical Data Fields {tbl-colwidths="\[35,35,30\]"}

Cancer types are reported using the OncoTree ontology originally developed at Memorial Sloan Kettering Cancer Center. The 13.0-public uses the OncoTree version oncotree_2021_11_02. The centers
participating in GENIE applied the OncoTree cancer types to the tested specimens in a variety of methods depending on center-specific workflows. A brief description of how the cancer type assignment process
for each center is specified in Table 9.

# Abbreviations and Acronym Glossary

For center abbreviations please see Table 1.

+------------------+-----------------------------------------------------------------+
| **Abbreviation** | **Full Term**                                                   |
+==================+=================================================================+
| AACR             | American Association for Cancer Research, Philadelphia, PA, USA |
+------------------+-----------------------------------------------------------------+
| CNA              | Copy number alterations                                         |
+------------------+-----------------------------------------------------------------+
| CNV              | Copy number variants                                            |
+------------------+-----------------------------------------------------------------+
| FFPE             | Formalin-fixed, paraffin-embedded                               |
+------------------+-----------------------------------------------------------------+
| GENIE            | Genomics, Evidence, Neoplasia, Information, Exchange            |
+------------------+-----------------------------------------------------------------+
| HIPAA            | Health Insurance Portability and Accountability Act             |
+------------------+-----------------------------------------------------------------+
| IRB              | Institutional Review Board                                      |
+------------------+-----------------------------------------------------------------+
| MAF              | Mutation annotation format                                      |
+------------------+-----------------------------------------------------------------+
| NAACCR           | North American Association of Central Cancer Registries         |
+------------------+-----------------------------------------------------------------+
| NGS              | Next-generation sequencing                                      |
+------------------+-----------------------------------------------------------------+
| PCR              | Polymerase chain reaction                                       |
+------------------+-----------------------------------------------------------------+
| SNP              | Single-nucleotide polymorphism                                  |
+------------------+-----------------------------------------------------------------+
| SNV              | Single-nucleotide variants                                      |
+------------------+-----------------------------------------------------------------+
| VCF              | Variant Call Format                                             |
+------------------+-----------------------------------------------------------------+
